---
title: "Figure 2 - Biological result improvement"
author: "Lucas Graybuck"
date: "5/29/2020"
output: html_document
---

## Setup

Load Packages
```{r}
quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}

quiet_library(H5weaver)
quiet_library(ATAComb)
quiet_library(uwot)
quiet_library(readr)
quiet_library(dplyr)
quiet_library(purrr)
quiet_library(ggplot2)
quiet_library(ggrastr)
quiet_library(cowplot)
options(stringsAsFactors = FALSE)
set.seed(3030)
```

Source directories
```{r}
filt_meta_dir <- "../../Data/X041_downsampled_200M/filtered_metadata/"
archr_res_dir <- "../../Data/X041_downsampled_200M/archr_results/"
window_mat_dir <- "../../Data/X041_downsampled_200M/window_20k_h5/"
```

Selected wells:  
X041-AP0C1W1, X041_aifi-nuclei-facs-unsorted, nuclear prep with standard 10x protocol  
X041-AP0C1W2, X041_aifi-nuclei-facs-sort-dd, nuclear prep with 10x protocol and FACS to remove debris and dead cells  
X041-AP0C1W3, X041_aifi-nuclei-facs-sort-neu-dd, nuclear prep with 10x protocol and FACS to remove neutrophils, debris, and dead cells  
X041-AP0C1W4, X041_aifi-perm-facs-unsorted, perm cell prep with no FACS  
X041-AP0C1W5, X041_aifi-perm-facs-sort-dd, perm cell prep with FACS to remove debris and dead cells
X041-AP0C1W6, X041_aifi-perm-facs-sort-neu-dd, perm cell prep with FACS to remove neutrophils, debris, and dead cells
```{r}
wells_to_process <- list(nuc_no_facs = "X041_aifi-nuclei-facs-unsorted",
                         nuc_dd_facs = "X041_aifi-nuclei-facs-sort-dd",
                         nuc_ndd_facs = "X041_aifi-nuclei-facs-sort-neu-dd",
                         pc_no_facs = "X041_aifi-perm-facs-unsorted",
                         pc_dd_facs = "X041_aifi-perm-facs-sort-dd",
                         pc_ndd_facs = "X041_aifi-perm-facs-sort-neu-dd")
well_labels <- c("Nuclei: Unsorted",
                 "Nuclei: No Dead/Debris",
                 "Nuclei: No Neutrophils",
                 "Perm: Unsorted",
                 "Perm: No Dead/Debris",
                 "Perm: No Neutrophils")
```

Locate files
```{r}
filt_meta_files <- map2(filt_meta_dir, wells_to_process, list.files, full.names = TRUE)
archr_res_files <- map2(archr_res_dir, wells_to_process, list.files, full.names = TRUE)
window_mat_files <- map2(window_mat_dir, wells_to_process, list.files, full.names = TRUE)
type_anno_file <- "../common/seurat_pbmc_type_anno.tsv"
```

```{r}
well_colors <- read.csv("../common/x041_well_colors.csv")
```


Display inputs
```{r}
print("Filtered Metadata Files:")
print(unlist(filt_meta_files))
```

```{r}
print("ArchR Results Files:")
print(unlist(archr_res_files))
```

```{r}
print("Peak Matrix Files:")
print(unlist(window_mat_files))
```

Read metadata files
```{r message=FALSE}
filt_meta_list <- map(filt_meta_files, 
                      read_csv)
```

Read ArchR results files
```{r message=FALSE}
archr_res_list <- map(archr_res_files, 
                      read_csv)
```

Read peak matrices
```{r}
window_mats <- map(window_mat_files,
                 read_h5_dgCMatrix)
```

### Filtering to remove doublets
ArchR results are filtered to remove doublet cells.
```{r message=FALSE}
type_anno <- read_tsv(type_anno_file) %>%
  filter(seurat_pbmc_type != "Platelets")

meta_list <- map2(filt_meta_list, archr_res_list,
                  function(filt_meta, archr_res) {
                    left_join(archr_res, filt_meta)
                  })
celltypes <- unique(unlist(map(meta_list, function(meta) unique(meta$seurat_pbmc_type))))

meta_list <- map(meta_list,
                  function(meta) {
                    left_join(meta, type_anno)
                  })
```

Filter peak data too
```{r}
window_mats <- map2(window_mats, meta_list,
                  function(window_mat, meta) {
                    window_mat[,meta$barcodes]
                  })
```


### Dimension reduction
```{r}
peak_lsi <- map(window_mats,
                atac_lsi)
peak_pcs <- map(peak_lsi,
                function(lsi) {
                  lsi$pcs[,2:30]
                })
```

```{r}
peak_umaps <- map(peak_pcs,
                 umap,
                 scale = TRUE,
                 min_dist = 0.2)
peak_umap_dfs <- map2(meta_list, peak_umaps,
                      function(filt_meta, peak_umap) {
                        peak_umap <- as.data.frame(peak_umap)
                        names(peak_umap) <- c("umap_1","umap_2")
                        cbind(filt_meta, peak_umap)
                      })
```

```{r}
peak_plots_umap_log_unique <- map2(peak_umap_dfs, well_labels,
                       function(peak_umap_df, well_label) {
                         ggplot(peak_umap_df) +
                           geom_point_rast(aes(x = umap_1,
                                               y = umap_2,
                                               color = peaks_frac)) +
                           scale_color_viridis_c("FRIP",
                                                 limits = c(0,1),
                                                 option = "magma") +
                           theme_bw(base_size = 7) +
                           theme(legend.position = "bottom",
                                 axis.text = element_text(color = "black"),
                                 axis.ticks = element_blank(),
                                 panel.grid.minor = element_blank(),
                                 panel.border = element_blank()) +
                           guides(color = guide_colourbar(title.position = "top",
                                                          barheight = unit(0.1, "in"))) +
                           ggtitle(well_label)
                       })
```


```{r fig.width = 7.5, fig.height = 2}
peak_plots_umap_frip_grid <- plot_grid(plotlist = peak_plots_umap_log_unique,
                                 ncol = 6)
peak_plots_umap_frip_grid
```

```{r}
ggsave("Figure_2A_umaps_frip.pdf",
       peak_plots_umap_frip_grid,
       width = 7.5, height = 2,
       useDingbats = FALSE)
```

```{r}
peak_plots_umap_celltype <- map2(peak_umap_dfs, well_labels,
                       function(peak_umap_df, well_label) {
                         ggplot(peak_umap_df) +
                           geom_point_rast(aes(x = umap_1,
                                               y = umap_2,
                                               color = seurat_pbmc_type_color)) +
                           theme_bw(base_size = 7) +
                           scale_color_identity() +
                           theme(legend.position = "bottom",
                                 axis.text = element_text(color = "black"),
                                 axis.ticks = element_blank(),
                                 panel.grid.minor = element_blank(),
                                 panel.border = element_blank()) +
                           ggtitle(well_label)
                       })
```

```{r}
peak_plots_umap_celltype_grid <- plot_grid(plotlist = peak_plots_umap_celltype,
                                 ncol = 6)
```

```{r fig.width = 7.5, fig.height = 1.6}
peak_plots_umap_celltype_grid
```

```{r fig.width = 1.5, fig.height = 1.75}
peak_plots_umap_celltype_legend <- ggplot(type_anno) +
  geom_tile(aes(x = 1,
                y = nrow(type_anno):1,
                fill = seurat_pbmc_type_color)) +
  geom_text(aes(x = 1.6,
                y = nrow(type_anno):1,
                label = seurat_pbmc_type),
            hjust = 0,
            size = 3) +
  scale_fill_identity() +
  xlim(0,8)+
  theme_void()
peak_plots_umap_celltype_legend
```

```{r}
ggsave("Figure_2A_umaps_celltype.pdf",
       peak_plots_umap_celltype_grid,
       width = 7.5, height = 1.6,
       useDingbats = FALSE)
ggsave("Figure_2A_umaps_celltype_legend.pdf",
       peak_plots_umap_celltype_legend,
       width = 1.5, height = 1.75,
       useDingbats = FALSE)
```

### Type fraction barplot
```{r}
flow_count_summary <- read.csv("../common/2020-07-21_IBD3_FACS_gating_counts.csv")
flow_count_summary <- flow_count_summary %>%
  dplyr::rename(n_cells = count,
                value = frac) %>%
  left_join(type_anno) %>%
  arrange(seurat_pbmc_type_id) %>%
  mutate(well_label = "Flow Cytometry",
         wpos = 7,
         vmax = cumsum(value),
         vmin = lag(vmax, default = 0))
```


```{r}
celltype_count_summary <- map2_dfr(meta_list, well_labels,
                         function(meta, well_label) {
                           count_list <- map(celltypes,
                                             function(celltype) {
                                               sum(meta$seurat_pbmc_type == celltype) 
                                             })
                           names(count_list) <- celltypes
                           cbind(data.frame(well_label = well_label,
                                            n_cells = nrow(meta)),
                                 as.data.frame(count_list))
                         })
celltype_count_summary
```

```{r}
celltype_frac_summary <- celltype_count_summary
for(i in 3:ncol(celltype_frac_summary)) {
  celltype_frac_summary[[i]] <- celltype_frac_summary[[i]] / celltype_frac_summary$n_cells
}
celltype_frac_summary
```

```{r}
celltype_frac_summary$wpos <- nrow(celltype_frac_summary):1
celltype_frac_summary_plot_df <- reshape2::melt(celltype_frac_summary,
                                  c("well_label","wpos","n_cells")) %>%
  rename(seurat_pbmc_type = variable) %>%
  left_join(type_anno) %>%
  arrange(well_label, seurat_pbmc_type_id) %>%
  group_by(well_label) %>%
  mutate(vmax = cumsum(value),
         vmin = lag(vmax, default = 0))

flow_count_summary <- flow_count_summary %>%
  select(names(celltype_frac_summary_plot_df))

celltype_frac_summary_plot_df <- rbind(celltype_frac_summary_plot_df, flow_count_summary)
```

RMSD calculations
```{r}
celltype_frac_mat <- as.matrix(celltype_frac_summary[,3:(ncol(celltype_frac_summary)-1)])
rownames(celltype_frac_mat) <- celltype_frac_summary$well_label

# Add zeroes for any cell type not detected in any of our conditions (e.g. T.CD8.Naive)
missing_types <- setdiff(flow_count_summary$seurat_pbmc_type, colnames(celltype_frac_mat))

missing_mat <- matrix(rep(0, length(missing_types) * nrow(celltype_frac_mat)),
                      ncol = length(missing_types))
colnames(missing_mat) <- missing_types
celltype_frac_mat <- cbind(celltype_frac_mat, missing_mat)
celltype_frac_mat <- celltype_frac_mat[,flow_count_summary$seurat_pbmc_type]

flow_frac_vec <- flow_count_summary$value

rmsd_values <- apply(celltype_frac_mat,
                     1,
                     function(atac_frac_vec) {
                       d <- flow_frac_vec - atac_frac_vec
                       sq_d <- d ^ 2
                       m_sq_d <- mean(sq_d)
                       round(sqrt(m_sq_d), 3)
                     })
rmsd_values
```


```{r}
y_labels <- celltype_frac_summary_plot_df %>%
  select(well_label, wpos) %>%
  unique()

vlines <- data.frame(x = seq(0,1,0.25))
celltype_frac_summary_plot <- ggplot() +
  geom_rect(data = celltype_frac_summary_plot_df,
            aes(xmin = vmin, xmax = vmax,
                ymin = wpos - 0.4, ymax = wpos + 0.4,
                fill = seurat_pbmc_type_color)) +
  geom_vline(data = vlines, 
             aes(xintercept = x),
             linetype = "dashed",
             size = 0.2,
             color = "black") +
  scale_fill_identity() +
  scale_x_continuous("Fraction of labeled cells") +
  scale_y_continuous(breaks = y_labels$wpos,
                     labels = y_labels$well_label) +
  guides(fill=guide_legend(nrow=2,byrow=TRUE)) +
  theme_bw(base_size = 7) +
  theme(axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "none")
```

```{r fig.width = 4, fig.height = 1.5}
celltype_frac_summary_plot
```

```{r}
ggsave("Figure_2B_celltype_fraction_barplot.pdf",
       celltype_frac_summary_plot,
       width = 4, height = 1.5,
       useDingbats = FALSE)
```


### Type transfer score distributions

Score dropoff coordinates
```{r}
score_drop_df <- map2_dfr(meta_list, well_labels,
                          function(meta, well_label) {
                            meta %>%
                              mutate(well_label = well_label) %>%
                              arrange(desc(seurat_pbmc_type_score)) %>%
                              mutate(rank = 1:n()) %>%
                              select(well_label, seurat_pbmc_type_score, rank) %>%
                              mutate(frac = rank / n())
                          })
```

Score dropoff plot
```{r fig.width = 3, fig.height = 1.5}

score_drop_plot <- ggplot(score_drop_df) +
  geom_line(aes(x = rank,
                y = seurat_pbmc_type_score,
                group = well_label,
                color = well_label),
            size = 0.5) +
  scale_x_continuous("Cells ranked by score") +
  scale_y_continuous("Seurat label transfer score") +
  scale_color_manual("",
                       breaks = well_colors$well_label,
                       values = well_colors$well_color) +
  theme_bw(base_size = 7)
score_drop_plot
```
```{r}
ggsave("Figure_2C_labeling_score_drop_plot.pdf",
       score_drop_plot,
       width = 3, height = 1.5,
       useDingbats = FALSE)
```

```{r}
sessionInfo()
```
