---
title: "Supplementary Figure 1 analysis"
author: "Lucas Graybuck"
date: "8/27/2020"
output: html_document
---

## Setup

Load Packages
```{r}
quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}

quiet_library(data.table)
quiet_library(readr)
quiet_library(dplyr)
quiet_library(purrr)
quiet_library(ggplot2)
quiet_library(ggrastr)
quiet_library(cowplot)
quiet_library(patchwork)
quiet_library(GenomicRanges)
options(stringsAsFactors = FALSE)
source("../common/granges_helpers.R")
```

Source directories
```{r}
full_meta_dir <- "../../Data/downsampled_125M/all_metadata/"
filt_meta_dir <- "../../Data/downsampled_125M/filtered_metadata/"
fragments_dir <- "../../Data/downsampled_125M/filtered_fragments/"
failed_fragments_dir <- "../../Data/downsampled_125M/failed_fragments/"
singlecell_dir <- "../../Data/downsampled_125M/singlecell/"
```

Selected wells:  

Nuclear Lysis protocols:
X024-AP0C1W1, Ficoll, Nuclei, 1x 10xNIB
X024-AP0C1W2, Ficoll, Nuclei, 0.25x 10xNIB
X024-AP0C1W3, Ficoll, Nuclei, 0.1x 10xNIB
X024-AP0C1W4, Ficoll, Nuclei, 1x ANIB

Permeabilization protocols:
X025-AP0C1W1, Ficoll, Perm Cells, 0.01% Digitonin
X025-AP0C1W2, Ficoll, Perm Cells, 0.05% Digitonin
X025-AP0C1W3, Ficoll, Perm Cells, 0.1% Digitonin
X025-AP0C1W4, Ficoll, Perm Cells, 0.2% Digitonin

Additional Washes:
X018-AP0C1W1, Lueka, Perm Cells, 0.01% Digitonin, Standard
X018-AP0C1W3, Lueka, Perm Cells, 0.01% Digitonin, 3 Washes Pre-perm
X018-AP0C1W4, Lueka, Perm Cells, 0.01% Digitonin, 3 Washes Pre-perm, 1 Wash Post-perm

Ficoll FACS vs Leukapheresis FACS:
X027-AP0C1W1, Ficoll, Perm Cells, No FACS
X027-AP0C1W3, Ficoll, Perm Cells, Neutrophil Depletion
B003-AP0C1W7, Lueka, Perm Cells, No FACS
B003-AP0C1W8, Leuka, Perm Cells, Neutrophil Depletion

Magnetic Bead purification:
X032-AP0C1W3, Ficoll, Perm Cells, No Beads
X032-AP0C1W4, Ficoll, Perm Cells, CD15 Depletion
X032-AP0C1W1, Leuka, Perm Cells, No Beads
X032-AP0C1W2, Leuka, Perm Cells, CD15 Depletion

```{r}
wells_to_process <- list("X024-AP0C1W1",
                         "X024-AP0C1W2",
                         "X024-AP0C1W3",
                         "X024-AP0C1W4",
                         "X025-AP0C1W1",
                         "X025-AP0C1W2",
                         "X025-AP0C1W3",
                         "X025-AP0C1W4",
                         # "X018-AP0C1W1",
                         # "X018-AP0C1W3",
                         # "X018-AP0C1W4",
                         "X027-AP0C1W1",
                         "X027-AP0C1W3",
                         "B003-AP0C1W7",
                         "B003-AP0C1W8",
                         "X032-AP0C1W3",
                         "X032-AP0C1W4",
                         "X032-AP0C1W1",
                         "X032-AP0C1W2")
well_labels <- c("Ficoll, Nuclei, 1x 10xNIB",
                 "Ficoll, Nuclei, 0.25x 10xNIB",
                 "Ficoll, Nuclei, 0.1x 10xNIB",
                 "Ficoll, Nuclei, 1x ANIB",
                 "Ficoll, Perm Cells, 0.01% Digitonin",
                 "Ficoll, Perm Cells, 0.05% Digitonin",
                 "Ficoll, Perm Cells, 0.1% Digitonin",
                 "Ficoll, Perm Cells, 0.2% Digitonin",
                 # "Lueka, Perm Cells, 0.01% Digitonin, Standard",
                 # "Lueka, Perm Cells, 0.01% Digitonin, 3 Washes Pre-perm",
                 # "Lueka, Perm Cells, 0.01% Digitonin, 3 Washes Pre-perm, 1 Wash Post-perm",
                 "Ficoll, Perm Cells, No FACS",
                 "Ficoll, Perm Cells, Neutrophil Depletion",
                 "Lueka, Perm Cells, No FACS",
                 "Leuka, Perm Cells, Neutrophil Depletion",
                 "Ficoll, Perm Cells, No Beads",
                 "Ficoll, Perm Cells, CD15 Depletion",
                 "Leuka, Perm Cells, No Beads",
                 "Leuka, Perm Cells, CD15 Depletion")
```

Locate files
```{r}
full_meta_files <- map2(full_meta_dir, wells_to_process, list.files, full.names = TRUE)
filt_meta_files <- map2(filt_meta_dir, wells_to_process, list.files, full.names = TRUE)
singlecell_files <- map2(singlecell_dir, wells_to_process, list.files, full.names = TRUE)
```

Display inputs
```{r}
print("Full Metadata Files:")
print(unlist(full_meta_files))
```

```{r}
print("Filtered Metadata Files:")
print(unlist(filt_meta_files))
```

```{r}
print("10x singlecell.csv Files:")
print(unlist(singlecell_files))
```


Read metadata files
```{r message=FALSE}
full_meta_list <- map(full_meta_files, read_csv)
filt_meta_list <- map(filt_meta_files, read_csv)
```


## Reads vs FRIP plots

Flag full metadata cells that pass or fail filtering
```{r}
full_meta_list <- map2(full_meta_list, filt_meta_list,
                       function(full, filt) {
                         full %>%
                           mutate(pass_fail = ifelse(barcodes %in% filt$barcodes,
                                                     "Pass", "Fail"))
                       })
```

Generate depth vs FRIP plots
```{r}
reads_vs_frip_plotlist <- map2(full_meta_list, well_labels,
                               function(meta, label) {
                                 ggplot(meta) +
                                   geom_point_rast(aes(x = n_unique,
                                                       y = peaks_frac,
                                                       color = pass_fail),
                                                   size = 0.5,
                                                   raster.dpi = 150) +
                                   scale_color_manual(breaks = c("Fail","Pass"),
                                                      values = c("#404040","skyblue")) +
                                   scale_x_log10(limits = c(8e2, 2e5),
                                                 breaks = c(1e3, 1e4, 1e5),
                                                 labels = c("1k", "10k", "100k"),
                                                 expand = c(0,0)) +
                                   xlab(bquote(log[10]("Unique Reads"))) +
                                   scale_y_continuous("",
                                                      limits = c(0, 1),
                                                      expand = c(0,0)) +
                                   theme_bw(base_size = 6) +
                                   theme(legend.position = "none",
                                         panel.grid.major = element_line(colour = "gray80",
                                                                         size = 0.2),
                                         panel.grid.minor = element_blank(),
                                         panel.border = element_blank(),
                                         axis.ticks = element_blank(),
                                         axis.text = element_text(colour = "black")) +
                                   ggtitle(label)
                               })
```

Preview depth vs FRIP plots
```{r fig.width = 6, fig.height = 7.25}
reads_vs_frip_plot_grid <- plot_grid(plotlist = reads_vs_frip_plotlist,
                                     ncol = 4)
reads_vs_frip_plot_grid
```

Save plots
```{r}
ggsave("Supp_Fig_1_reads_vs_frip_plot.pdf",
       width = 6, height = 7.25,
       useDingbats = FALSE)
```


## Fragment size distributions

Fragments in barcodes that pass QC
```{r message=FALSE}
passed_fragment_length_distributions <- map(fragment_files,
                                            function(fragment_file) {
                                              print(paste("Tabulating widths for",fragment_file))
                                              
                                              frags_df <- read_tsv(fragment_file, 
                                                                   col_names = c("seqnames","start","end","barcodes","count"))
                                              frags_df$width <- frags_df$end - frags_df$start
                                              frag_width_table <- frags_df %>%
                                                group_by(width) %>%
                                                summarise(count = n()) %>%
                                                ungroup()
                                              frag_width_table$frac <- frag_width_table$count / sum(frag_width_table$count)
                                              frag_width_table
                                            })
```

Fragments in barcodes that fail QC
```{r message=FALSE}
failed_fragment_length_distributions <- map(failed_fragment_files,
                                            function(fragment_file) {
                                              print(paste("Tabulating widths for",fragment_file))
                                              
                                              frags_df <- read_tsv(fragment_file, 
                                                                   col_names = c("seqnames","start","end","barcodes","count"))
                                              frags_df$width <- frags_df$end - frags_df$start
                                              frag_width_table <- frags_df %>%
                                                group_by(width) %>%
                                                summarise(count = n()) %>%
                                                ungroup()
                                              frag_width_table$frac <- frag_width_table$count / sum(frag_width_table$count)
                                              frag_width_table
                                            })
```
Build Plots
```{r}
frag_length_plots <- pmap(list(passed_fragment_length_distributions,
                               failed_fragment_length_distributions,
                               well_labels),
                          function(passed_fragment_length_distribution,
                                   failed_fragment_length_distributions,
                                   well_label) {
                            well_color <- well_colors$well_color[well_colors$well_label == well_label]
                            
                            passed_fragment_length_distribution$pass_fail <- "Pass"
                            failed_fragment_length_distributions$pass_fail <- "Fail"
                            dist_df <- rbind(passed_fragment_length_distribution,
                                             failed_fragment_length_distributions)
                            ggplot(dist_df) +
                              geom_line(aes(x = width,
                                            y = frac,
                                            color = pass_fail)) +
                              scale_color_manual(breaks = c("Fail","Pass"),
                                                 values = c("#404040",well_color)) +
                              scale_x_continuous("Fragment length",
                                                 limits = c(0,500)) +
                              scale_y_continuous("Frac. reads") +
                              theme_bw(base_size = 7) +
                              theme(legend.position = "none",
                                    panel.grid.major = element_line(colour = "gray80",
                                                                    size = 0.2),
                                    panel.grid.minor = element_blank(),
                                    panel.border = element_blank(),
                                    axis.ticks = element_blank(),
                                    axis.text = element_text(colour = "black"))
                          })
```

```{r}
fragment_length_plot_grid <- plot_grid(plotlist = frag_length_plots, ncol = 6)
```

```{r fig.width = 7.5, fig.height = 0.8}
fragment_length_plot_grid
```

```{r}
ggsave("Figure_1B_fragment_length_plot.pdf",
       width = 7.5, height = 0.8,
       useDingbats = FALSE)
```


```{r}
sessionInfo()
```

