---
title: "Fragment size analysis"
author: "Lucas Graybuck"
date: "5/27/2020"
output: html_document
---

## Setup

Load Packages
```{r}
quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}

quiet_library(data.table)
quiet_library(dplyr)
quiet_library(purrr)
quiet_library(ggplot2)
quiet_library(ggrastr)
quiet_library(cowplot)
quiet_library(patchwork)
quiet_library(GenomicRanges)
options(stringsAsFactors = FALSE)
source("../common/granges_helpers.R")
```

Source directories
```{r}
full_meta_dir <- "../../Data/X041_downsampled_200M/all_metadata/"
filt_meta_dir <- "../../Data/X041_downsampled_200M/filtered_metadata/"
fragments_dir <- "../../Data/X041_downsampled_200M/filtered_fragments/"
failed_fragments_dir <- "../../Data/X041_downsampled_200M/failed_fragments/"
well_colors <- read.csv("../common/x041_well_colors.csv")
```

Selected wells:  
X041-AP0C1W1, X041_aifi-nuclei-facs-unsorted, nuclear prep with standard 10x protocol  
X041-AP0C1W2, X041_aifi-nuclei-facs-sort-dd, nuclear prep with 10x protocol and FACS to remove debris and dead cells  
X041-AP0C1W3, X041_aifi-nuclei-facs-sort-neu-dd, nuclear prep with 10x protocol and FACS to remove neutrophils, debris, and dead cells  
X041-AP0C1W4, X041_aifi-perm-facs-unsorted, perm cell prep with no FACS  
X041-AP0C1W5, X041_aifi-perm-facs-sort-dd, perm cell prep with FACS to remove debris and dead cells
X041-AP0C1W6, X041_aifi-perm-facs-sort-neu-dd, perm cell prep with FACS to remove neutrophils, debris, and dead cells
```{r}
wells_to_process <- list(nuc_no_facs = "X041_aifi-nuclei-facs-unsorted",
                         nuc_dd_facs = "X041_aifi-nuclei-facs-sort-dd",
                         nuc_ndd_facs = "X041_aifi-nuclei-facs-sort-neu-dd",
                         pc_no_facs = "X041_aifi-perm-facs-unsorted",
                         pc_dd_facs = "X041_aifi-perm-facs-sort-dd",
                         pc_ndd_facs = "X041_aifi-perm-facs-sort-neu-dd")
well_labels <- c("Nuclei: Unsorted",
                 "Nuclei: No Dead/Debris",
                 "Nuclei: No Neutrophils",
                 "Perm: Unsorted",
                 "Perm: No Dead/Debris",
                 "Perm: No Neutrophils")
```

Locate files
```{r}
full_meta_files <- map2(full_meta_dir, wells_to_process, list.files, full.names = TRUE)
filt_meta_files <- map2(filt_meta_dir, wells_to_process, list.files, full.names = TRUE)
fragment_files <- map2(fragments_dir, wells_to_process, list.files, full.names = TRUE)
fragment_files <- map(fragment_files, function(filenames) {filenames[!grepl(".tbi$", filenames)]})
failed_fragment_files <- map2(failed_fragments_dir, wells_to_process, list.files, full.names = TRUE)
failed_fragment_files <- map(failed_fragment_files, function(filenames) {filenames[!grepl(".tbi$", filenames)]})
```

Display inputs
```{r}
print("Full Metadata Files:")
print(unlist(full_meta_files))
```

```{r}
print("Filtered Metadata Files:")
print(unlist(filt_meta_files))
```
```{r}
print("Filtered Fragment Files:")
print(unlist(fragment_files))
```
```{r}
print("Failed Fragment Files:")
print(unlist(failed_fragment_files))
```

Read Peaks
```{r}
peaks_file <- system.file("reference/hg38_peaks_gr.rds", package = "ATAComb")
peaks_gr <- readRDS(peaks_file)
```


Read metadata files
```{r message=FALSE}
full_meta_list <- map(full_meta_files, fread)
filt_meta_list <- map(filt_meta_files, fread)
```


## Reads vs FRIP plots

Flag full metadata cells that pass or fail filtering
```{r}
full_meta_list <- map2(full_meta_list, filt_meta_list,
                       function(full, filt) {
                         full %>%
                           mutate(pass_fail = ifelse(barcodes %in% filt$barcodes,
                                                     "Pass", "Fail"))
                       })
```

## Read fragments
```{r}
pass_fragment_list <- map(fragment_files,
                          fread,
                          col.names = c("seqnames","start","end","barcodes","count"))
names(pass_fragment_list) <- well_labels
fail_fragment_list <- map(failed_fragment_files,
                          fread,
                          col.names = c("seqnames","start","end","barcodes","count"))
names(fail_fragment_list) <- well_labels
```
## Convert to GRanges
```{r}
pass_gr_list <- map(pass_fragment_list,
                    function(frags) {
                      GRanges(seqnames = frags$seqnames,
                              IRanges(start = frags$start,
                                      end = frags$end),
                              barcodes = frags$barcodes)
                    })
fail_gr_list <- map(fail_fragment_list,
                    function(frags) {
                      GRanges(seqnames = frags$seqnames,
                              IRanges(start = frags$start,
                                      end = frags$end),
                              barcodes = frags$barcodes)
                    })
```

## Filter for peak overlaps
```{r}
pass_peaks_gr <- map(pass_gr_list,
                     filter_gr,
                     peaks_gr,
                     mode = "keep")
pass_nonpeaks_gr <- map(pass_gr_list,
                        filter_gr,
                        peaks_gr,
                        mode = "remove")

fail_peaks_gr <- map(fail_gr_list,
                     filter_gr,
                     peaks_gr,
                     mode = "keep")
fail_nonpeaks_gr <- map(fail_gr_list,
                        filter_gr,
                        peaks_gr,
                        mode = "remove")
```


## Fragment size distributions

```{r}
gr_width_summary <- function(gr, wl, pf, loc) {
  df <- as.data.frame(table(width(gr)))
  names(df) <- c("width", "count")
  df$width <- as.numeric(as.character(df$width))
  df$frac <- df$count / sum(df$count)
  df$well_label <- wl
  df$pass_fail <- pf
  df$loc <- loc
  df
}

pass_peaks_sizes <-map2(pass_peaks_gr, well_labels,
                        gr_width_summary,
                        pf = "Pass", loc = "peaks")
pass_nonpeaks_sizes <-map2(pass_nonpeaks_gr, well_labels,
                           gr_width_summary,
                           pf = "Pass", loc = "nonpeaks")
fail_peaks_sizes <-map2(fail_peaks_gr, well_labels,
                        gr_width_summary,
                        pf = "Fail", loc = "peaks")
fail_nonpeaks_sizes <-map2(fail_nonpeaks_gr, well_labels,
                           gr_width_summary,
                           pf = "Fail", loc = "nonpeaks")
```

Build Plots
```{r}
frag_length_plots <- pmap(list(pass_peaks_sizes, pass_nonpeaks_sizes,
                               fail_peaks_sizes, fail_nonpeaks_sizes,
                               well_labels),
                          function(pp, pnp, fp, fnp, well_label) {
                            dfs <- list(pp, pnp, fp, fnp)
                            clrs <- c("#27AAE1","#27AAE1","#BE1E2D","#BE1E2D")
                            map2(dfs, clrs,
                                function(df, clr) {
                                  ggplot(df) +
                                    geom_line(aes(x = width,
                                                  y = frac),
                                                  color = clr) +
                                    scale_x_continuous("Fragment length",
                                                       limits = c(0,500)) +
                                    scale_y_continuous("Frac. reads") +
                                    theme_bw(base_size = 7) +
                                    theme(legend.position = "none",
                                          panel.grid.major = element_line(colour = "gray80",
                                                                          size = 0.2),
                                          panel.grid.minor = element_blank(),
                                          panel.border = element_blank(),
                                          axis.ticks = element_blank(),
                                          axis.text = element_text(colour = "black")) +
                                    ggtitle(paste(df$well_label, df$pass_fail, df$loc))
                                })
                          })
frag_length_plots <- unlist(frag_length_plots, recursive = FALSE)
```

```{r}
fragment_length_plot_grid <- plot_grid(plotlist = frag_length_plots, ncol = 4)
```

```{r fig.width = 7.5, fig.height = 6}
fragment_length_plot_grid
```

### Frag category stats

Fraction Summary
```{r}
mn_cut <- 120
dn_cut <- 280

all_sizes <- list(pass_peaks_sizes, pass_nonpeaks_sizes,
                  fail_peaks_sizes, fail_nonpeaks_sizes)
all_sizes <- lapply(all_sizes,
                    function(l) { do.call(rbind, l) })
all_sizes <- do.call(rbind, all_sizes)

all_sizes <- all_sizes %>%
  mutate(cat = case_when(width <= 120 ~ "0_open",
                         width > 120 & width <= 280 ~ "1_mono",
                         width > 280 ~ "2_dinuc"))

size_summary <- all_sizes %>%
  group_by(well_label, pass_fail, loc, cat) %>%
  summarise(n_frags = sum(count)) %>%
  group_by(well_label, pass_fail, loc) %>%
  mutate(frac_frags = n_frags / sum(n_frags)) %>%
  arrange(well_label, desc(pass_fail), desc(loc), cat)

size_summary
```

Simple fraction barplots
```{r fig.width = 4, fig.height = 4}
ggplot(size_summary) +
  geom_bar(aes(y = paste(well_label, pass_fail, loc),
               x = frac_frags,
               fill = cat),
           stat = "identity",
           position = position_stack(reverse = TRUE)) +
  scale_fill_manual("Category",
                    breaks = c("0_open", "1_mono","2_dinuc"),
                    labels = c("Open","Mono","Di+"),
                    values = c("gray60","gray40","gray20")) +
  scale_y_discrete("") +
  scale_x_continuous("Fraction of Unique Fragments",
                     expand = c(0,0)) +
  theme_bw(base_size = 7) +
  theme(legend.position = "right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(colour = "black"))
```


Chi-square tests
```{r}
chisq_mats <- map(well_labels,
                  function(wl) {
                    well_summary <- size_summary %>%
                      filter(well_label == wl)
                    well_mat <- matrix(well_summary$n_frags,
                           ncol = 2, nrow = 3)
                    rownames(well_mat) <- c("0_open","1_mono","2_dinuc")
                    colnames(well_mat) <- c("Fail","Pass")
                    well_mat
                  })

chisq_res <- map(chisq_mats,
                 chisq.test)
```

### Pileups based on size and pass-fail/peaks categories

```{r}
pass_fail_gr_list <- list(pass_peaks = pass_peaks_gr,
                          pass_nonpeaks = pass_nonpeaks_gr,
                          fail_peaks = fail_peaks_gr,
                          fail_nonpeaks = fail_nonpeaks_gr)

open_pass_fail_gr_list <- map(pass_fail_gr_list,
                              function(gr_list) {
                                map(gr_list,
                                    function(gr) {
                                      gr[width(gr) <= mn_cut]
                                    })
                              })

mono_pass_fail_gr_list <- map(pass_fail_gr_list,
                              function(gr_list) {
                                map(gr_list,
                                    function(gr) {
                                      gr[width(gr) > mn_cut & width(gr) <= dn_cut]
                                    })
                              })

long_pass_fail_gr_list <- map(pass_fail_gr_list,
                              function(gr_list) {
                                map(gr_list,
                                    function(gr) {
                                      gr[width(gr) > dn_cut]
                                    })
                              })
```

```{r}
tss_2kb_file <- system.file("reference/hg38_tss_gr.rds", package = "ATAComb")
tss_2kb <- readRDS(tss_2kb_file)
```

```{r}
all_regions <- unlist(list(open = open_pass_fail_gr_list,
                           mono = mono_pass_fail_gr_list,
                           long = long_pass_fail_gr_list))

tss_regions <- map(all_regions,
                   filter_gr,
                   target_gr = tss_2kb,
                   mode = "keep")

if(!file.exists("Figure_1_category_tss_pileups_list.rds")) {
  tss_pileups <- map(tss_regions,
                     function(fragments) {
                       pileup_fragments(query_gr = fragments,
                                        target_gr = tss_2kb,
                                        target_width = 4000)
                     })
  saveRDS(tss_pileups,
          "Figure_1_category_tss_pileups_list.rds")
} else {
  tss_pileups <- readRDS("Figure_1_category_tss_pileups_list.rds")
}
```

```{r}
tss_pileup_df <- map2_dfr(tss_pileups, names(tss_pileups),
                          function(tss_pileup, pile_name) {
                            name_split <- strsplit(pile_name, "\\.")[[1]]
                            tss_pileup$size_cat <- name_split[1]
                            tss_pileup$pass_fail <- sub("_.+","",name_split[2])
                            tss_pileup$peak_cat <- sub(".+_","",name_split[2])
                            tss_pileup$well_label <- name_split[3]
                            tss_pileup$frac_max <- tss_pileup$cov / max(tss_pileup$cov)
                            tss_pileup
                          })
```

# Arrange factor levels so plots are arranged appropriately
```{r}
tss_pileup_df$size_cat <- factor(tss_pileup_df$size_cat,
                                 levels = c("open","mono","long"))
tss_pileup_df$well_label <- factor(tss_pileup_df$well_label,
                                   levels = c("Nuclei: Unsorted","Nuclei: No Dead/Debris","Nuclei: No Neutrophils",
                                              "Perm: Unsorted","Perm: No Dead/Debris","Perm: No Neutrophils"))

well_colors$well_label <- factor(well_colors$well_label,
                                   levels = c("Nuclei: Unsorted","Nuclei: No Dead/Debris","Nuclei: No Neutrophils",
                                              "Perm: Unsorted","Perm: No Dead/Debris","Perm: No Neutrophils"))

filtered_tss_pileup_df <- tss_pileup_df %>%
  group_by(well_label, size_cat, pass_fail, pos) %>%
  summarise(cov = sum(cov)) %>%
  group_by(well_label, size_cat, pass_fail) %>%
  mutate(frac_max = cov / max(cov)) %>%
  left_join(well_colors) %>%
  mutate(plot_color = ifelse(pass_fail == "pass", well_color, "#404040"))
```


```{r fig.width = 5, fig.height = 5}
tss_fragment_coverage_plot <- ggplot(data = filtered_tss_pileup_df) +
  geom_vline(aes(xintercept = 2000),
             size = 0.5,
             color = "black") +
  geom_line(aes(x = pos,
                y = cov,
                group = paste(size_cat, pass_fail, well_label),
                color = plot_color),
            size = 0.5) +
  scale_color_identity() +
  facet_grid(rows = vars(well_label),
             cols = vars(size_cat),
             scales = "free_y") +
  scale_y_continuous("Total fragment coverage",
                     limits = c(0, 4e6)) +
  scale_x_continuous("Position Relative to TSS",
                     limits = c(1250, 2750),
                     breaks = seq(1250, 2750, 250),
                     labels = seq(-750, 750, 250)) +
  theme_bw(base_size = 7) +
  theme(panel.grid.major = element_line(colour = "gray80",
                                        size = 0.2),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(colour = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3))

tss_fragment_coverage_plot
```

```{r}
ggsave("Supp_Fig_4_unscaled_tss_fragment_coverage_plot.pdf",
       tss_fragment_coverage_plot,
       width = 4, height = 4,
       useDingbats = FALSE)
```



```{r fig.width = 5, fig.height = 5}
max_value_df <- filtered_tss_pileup_df %>%
  group_by(well_label, size_cat, pass_fail, plot_color) %>%
  summarise(max_cov = max(cov)) %>%
  mutate(max_cov_label = formatC(max_cov, format = "e", digits = 2))

pass_max_df <- max_value_df %>%
  filter(pass_fail == "Pass")
fail_max_df <- max_value_df %>%
  filter(pass_fail == "Fail")

scaled_tss_fragment_coverage_plot <- ggplot(data = filtered_tss_pileup_df) +
  geom_vline(aes(xintercept = 2000),
             size = 0.5,
             color = "black") +
  # geom_text(data = pass_max_df,
  #           aes(x = 2250,
  #               y = 0.9,
  #               label = paste0("max: max_cov_label")),
  #           color = " skyblue") +
  geom_line(aes(x = pos,
                y = frac_max,
                group = paste(size_cat, pass_fail, well_label),
                color = plot_color),
            size = 0.5) +
  scale_color_identity() +
  facet_grid(rows = vars(well_label),
             cols = vars(size_cat),
             scales = "free_y") +
  ylab("Fraction of max coverage") +
  scale_x_continuous("Pos. Relative to TSS",
                     limits = c(1250, 2750),
                     breaks = seq(1250, 2750, 250),
                     labels = seq(-750, 750, 250)) +
  theme_bw(base_size = 7) +
  theme(panel.grid.major = element_line(colour = "gray80",
                                        size = 0.2),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(colour = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3))

scaled_tss_fragment_coverage_plot
```

```{r}
ggsave("Supp_Fig_4_scaled_tss_fragment_coverage_plot.pdf",
       scaled_tss_fragment_coverage_plot,
       width = 4, height = 4,
       useDingbats = FALSE)
```



```{r}
sessionInfo()
```
