---
title: "QC Stats summary assembly"
author: "Lucas Graybuck"
date: "7/27/2020"
output: html_document
---

## Setup

Load Packages
```{r}
quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}

quiet_library(H5weaver)
quiet_library(readr)
quiet_library(dplyr)
quiet_library(purrr)
quiet_library(ggplot2)
```

Source directories
```{r}
full_meta_dir <- "../../Data/X041_downsampled_200M/all_metadata/"
filt_meta_dir <- "../../Data/X041_downsampled_200M/filtered_metadata/"
singlecell_dir <- "../../Data/X041_downsampled_200M/singlecell/"

peaks_mat_dir <- "../../Data/X041_downsampled_200M/peaks_h5/"
tss_mat_dir <- "../../Data/X041_downsampled_200M/tss_h5/"
```

Selected wells:  
X041-AP0C1W1, X041_aifi-nuclei-facs-unsorted, nuclear prep with standard 10x protocol  
X041-AP0C1W2, X041_aifi-nuclei-facs-sort-dd, nuclear prep with 10x protocol and FACS to remove debris and dead cells  
X041-AP0C1W3, X041_aifi-nuclei-facs-sort-neu-dd, nuclear prep with 10x protocol and FACS to remove neutrophils, debris, and dead cells  
X041-AP0C1W4, X041_aifi-perm-facs-unsorted, perm cell prep with no FACS  
X041-AP0C1W5, X041_aifi-perm-facs-sort-dd, perm cell prep with FACS to remove debris and dead cells
X041-AP0C1W6, X041_aifi-perm-facs-sort-neu-dd, perm cell prep with FACS to remove neutrophils, debris, and dead cells
```{r}
wells_to_process <- list(nuc_no_facs = "X041_aifi-nuclei-facs-unsorted",
                         nuc_dd_facs = "X041_aifi-nuclei-facs-sort-dd",
                         nuc_ndd_facs = "X041_aifi-nuclei-facs-sort-neu-dd",
                         pc_no_facs = "X041_aifi-perm-facs-unsorted",
                         pc_dd_facs = "X041_aifi-perm-facs-sort-dd",
                         pc_ndd_facs = "X041_aifi-perm-facs-sort-neu-dd")
well_labels <- c("Nuclei: Unsorted",
                 "Nuclei: No Dead/Debris",
                 "Nuclei: No Neutrophils",
                 "Perm: Unsorted",
                 "Perm: No Dead/Debris",
                 "Perm: No Neutrophils")
well_colors <- read.csv("../common/x041_well_colors.csv")
```


Locate files
```{r}
full_meta_files <- map2(full_meta_dir, wells_to_process, list.files, full.names = TRUE)
filt_meta_files <- map2(filt_meta_dir, wells_to_process, list.files, full.names = TRUE)
singlecell_files <- map2(singlecell_dir, wells_to_process, list.files, full.names = TRUE)
peaks_mat_files <- map2(peaks_mat_dir, wells_to_process, list.files, full.names = TRUE)
tss_mat_files <- map2(tss_mat_dir, wells_to_process, list.files, full.names = TRUE)
```

Display inputs
```{r}
print("Full Metadata Files:")
print(unlist(full_meta_files))
```

```{r}
print("Filtered Metadata Files:")
print(unlist(filt_meta_files))
```

```{r}
print("10x singlecell.csv Files:")
print(unlist(singlecell_files))
```

```{r}
print("Peak H5 Matrix Files:")
print(unlist(peaks_mat_files))
```

```{r}
print("TSS H5 Matrix Files:")
print(unlist(tss_mat_files))
```

Read metadata files
```{r message=FALSE}
full_meta_list <- map(full_meta_files, read_csv)
filt_meta_list <- map(filt_meta_files, read_csv)
singlecell_list <- map(singlecell_files, read_csv)
```

```{r}
filt_meta <- do.call(rbind, filt_meta_list) %>%
  mutate(frac_mito = n_mito / n_fragments) %>%
  mutate(frac_tss = tss_count / n_unique) %>%
  mutate(frac_peaks = peaks_count / n_unique)
```

Count Features for each dataset
```{r}
peak_counts <- map_dfr(peaks_mat_files,
                   function(peak_mat_file) {
                     print(paste("Processing", peak_mat_file))
                     peak_mat <- read_h5_dgCMatrix(peak_mat_file)
                     n_peak_hits <- colSums(peak_mat > 0)
                     data.frame(barcodes = colnames(peak_mat),
                                n_peak_hits = n_peak_hits,
                                frac_peak_hits = n_peak_hits / nrow(peak_mat))
                   })
tss_counts <- map_dfr(tss_mat_files,
                   function(tss_mat_file) {
                     print(paste("Processing", tss_mat_file))
                     tss_mat <- read_h5_dgCMatrix(tss_mat_file)
                     n_tss_hits <- colSums(tss_mat > 0)
                     data.frame(barcodes = colnames(tss_mat),
                                n_tss_hits = n_tss_hits,
                                frac_tss_hits = n_tss_hits / nrow(tss_mat))
                   })
```


```{r}
well_summary <- filt_meta %>%
  left_join(peak_counts) %>%
  left_join(tss_counts) %>%
  group_by(pbmc_sample_id) %>%
  summarise(n_cells = n(),
            med_n_fragments = median(n_fragments),
            med_n_mito = median(n_mito),
            med_frac_mito = median(n_mito / n_fragments),
            med_n_unique = median(n_unique),
            med_frac_unique = median(n_unique / n_fragments),
            med_n_tss_frag = median(tss_count),
            med_frac_tss_frag = median(tss_count / n_unique),
            med_n_tss_hits = median(n_tss_hits),
            med_frac_tss_hits = median(frac_tss_hits),
            med_n_peaks_frag = median(peaks_count),
            med_frac_peaks_frag = median(peaks_count / n_unique),
            med_n_peaks_hits = median(n_peak_hits),
            med_frac_peaks_hits = median(frac_peak_hits)) %>%
  mutate_at(vars(starts_with("med_frac")), round, digits = 3)

well_summary
```
```{r}
write.csv(well_summary,
          "X041_well_qc_stats.csv",
          row.names = FALSE)
```


Distribution plots for supplement:
```{r}
qc_violin_plot(filt_meta,
               category_x = "pbmc_sample_id",
               name_x = "Sample",
               column_y = "frac_mito",
               name_y = "Mito. Fraction",
               log_y = FALSE)
```

Distribution plots for supplement:
```{r}
qc_violin_plot(filt_meta,
               category_x = "pbmc_sample_id",
               name_x = "Sample",
               column_y = "n_unique",
               name_y = "N Unique Reads",
               log_y = TRUE)
```

```{r}
qc_violin_plot(filt_meta,
               category_x = "pbmc_sample_id",
               name_x = "Sample",
               column_y = "frac_tss",
               name_y = "FRITSS",
               log_y = FALSE)
```

```{r}
qc_violin_plot(filt_meta,
               category_x = "pbmc_sample_id",
               name_x = "Sample",
               column_y = "frac_peaks",
               name_y = "FRIP",
               log_y = FALSE)
```

