---
title: "ArchR Doublet and Labeling analysis"
author: 
 - Lucas Graybuck
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: true
params:
  in_frag: NULL
  in_sample: NULL
  genome: NULL
  n_threads: NULL
  out_dir: NULL
---

<a id="contents"></a>

## Contents

- [Session Preparation](#session_preparation)
- [Load References](#load_refs)

#### [ArchR Analysis](#archr_setup)
- [Cell Barcode QC](#archr_doublets)
- [Fragment Metrics](#archr_label)
- [Saturation Metrics](#archr_output)

[Write QC JSON](#json_out)

#### [Session Info](#session_info)

<a id="data_processing"></a>

## Data Processing

<a id="session_preparation"></a>

### Session Preparation

#### Load libraries:
```{r Load Libraries}
start_time <- Sys.time()

quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}
quiet_library(H5weaver)
quiet_library(ArchR)
quiet_library(data.table)
quiet_library(jsonlite)
options(stringsAsFactors = FALSE)
```

Declaring start
```{r Declare start}
stm("Starting ArchR analysis")
```

#### Argument parsing
```{r Parse arguments}
if(is.null(params$in_frag)) {
  in_frag <- system.file("testdata/outs/fragments.tsv.gz",
                         package = "ATAComb")
  in_sample <- "X000_PB000000"
  genome <- "hg38"
  out_dir <- tempdir()
  n_threads <- parallel::detectCores()
} else {
  in_frag <- params$in_frag
  in_sample <- params$in_sample
  genome <- params$genome
  out_dir <- params$out_dir
  n_threads <- as.numeric(params$n_threads)
}

# Check Genome
if(!genome %in% c("hg38","hg19")) {
  stm(paste0("Genome must be 'hg38' or 'hg19'. Genome provided: ", genome))
  stop()
}

if(is.null(n_threads)) {
  n_threads <- parallel::detectCores()
}

temp_dir <- file.path(out_dir, "temp")

stm(paste0("IN  fragments.tsv.gz      : ", in_frag))
stm(paste0("IN  Sample ID             : ", in_sample))
stm(paste0("IN  Genome                : ", genome))
stm(paste0("PAR N Threads             : ", n_threads))
stm(paste0("OUT Directory             : ", out_dir))
```

#### Input Parameters
```{r Print Arguments}
print(c(
  paste0("IN  fragments.tsv.gz      : ", in_frag),
  paste0("IN  Sample ID             : ", in_sample),
  paste0("IN  Genome                : ", genome),
  paste0("PAR N Threads             : ", n_threads),
  paste0("OUT Directory             : ", out_dir)
))
```

#### Check Input Files
```{r Check Inputs}
if(!file.exists(in_frag)) {
  stm(paste0("ERROR: Cannot find IN fragments.tsv.gz:", in_frag))
  stop()
}
```

#### Create out directory if missing
```{r Create Out Dir}
if(!dir.exists(out_dir)) {
  stm(paste0("Creating Output Directory: ",out_dir))
  dir.create(out_dir, 
             recursive = TRUE)
}
```

[Return to Contents](#contents)

<a id="load_refs"></a>

#### Load reference dataset
```{r}
reference_source <- system.file("reference/seurat_rna_pbmc_ref.h5",
                                package = "ATAComb")
file.copy(reference_source, file.path(out_dir,"seurat_pbmc_ref.h5"))
seurat_reference <- file.path(out_dir, "seurat_pbmc_ref.h5")
rna_so <- H5weaver::read_h5_seurat(seurat_reference)
file.remove(seurat_reference)
```

[Return to Contents](#contents)

<a id="archr_setup"></a>

#### Set ArchR parameters
```{r setup archr}
stm("Setting ArchR parameters")
set.seed(3030)

addArchRThreads(n_threads)
addArchRGenome(genome)
```

[Return to Contents](#contents)

<a id="archr_doublets"></a>

### ArchR Doublet process

#### Assemble Arrow from fragments
```{r}
stm(paste0("Reading fragments", in_frag))

arrowFiles <- createArrowFiles(inputFiles = in_frag,
                               sampleNames = in_sample,
                               filterTSS = 4,
                               filterFrags = 1000,
                               addTileMat = TRUE,
                               addGeneScoreMat = TRUE)
```

#### Score doublets
```{r Generate IDs}
stm("Scoring doublets")

doubScores <- addDoubletScores(input = arrowFiles,
                               k = 20,
                               knnMethod = "UMAP")
```

#### Set up ArchR Project
```{r}
stm("Building ArchR Project")
proj <- ArchRProject(ArrowFiles = arrowFiles,
                     outputDirectory = temp_dir)
```

#### Filter doublets based on standard cutoff
```{r}
stm("Filtering based on doublet cutoff")

proj <- filterDoublets(ArchRProj = proj,
                       cutEnrich = 1)
```

[Return to Contents](#contents)

<a id="archr_label"></a>

### ArchR Label transfer

#### Reduce ATAC dims with LSI and add clusters
```{r}
proj <- addIterativeLSI(ArchRProj = proj)
```

```{r}
proj <- addClusters(input = proj,
                    reducedDims = "IterativeLSI")
```


#### Run Unconstrained transfer
```{r}
outDir <- temp_dir
proj <- addGeneIntegrationMatrix(ArchRProj = proj,
                                 useMatrix = "GeneScoreMatrix",
                                 matrixName = "GeneIntegrationMatrix",
                                 reducedDims = "IterativeLSI",
                                 seRNA = rna_so,
                                 groupRNA = "celltype",
                                 nameCell = "un_pred_type",
                                 nameGroup = "un_pred_group",
                                 nameScore = "un_pred_score",
                                 force = TRUE
)
```

#### Unconstrained confusion matrix
```{r}
un_conf_mat <- as.matrix(confusionMatrix(proj$Clusters, proj$un_pred_group))
heatmap(un_conf_mat)
```

#### Partition for constrained transfer
```{r}
rna_types <- unique(rna_so@meta.data$celltype)

rna_patterns <- list(b_pdc = "^B|Plasmacytoid",
                     mono_dc = "^Mono|Myeloid",
                     tnk_cell = "^T|NK")

rna_clusters <- lapply(rna_patterns,
                       function(x) {
                           rna_types[grepl(x, rna_types)]
                       })

rna_cells <- lapply(rna_clusters,
                   function(x) {
                       colnames(rna_so)[rna_so@meta.data$celltype %in% x]
                   })
```

```{r}
preClust <- colnames(un_conf_mat)[apply(un_conf_mat, 1 , which.max)]

atac_clusters <- lapply(rna_clusters,
                        function(x) {
                            rownames(un_conf_mat)[preClust %in% x]
                        })

atac_cells <- lapply(atac_clusters,
                    function(x) {
                        proj$cellNames[proj$Clusters %in% x]
                    })

groupList <- lapply(names(atac_cells),
                   function(x) {
                       SimpleList(ATAC = atac_cells[[x]],
                                  RNA = rna_cells[[x]])
                   })
groupList <- SimpleList(groupList)
```

#### Constrained integration
```{r}
proj <- addGeneIntegrationMatrix(ArchRProj = proj,
                                 useMatrix = "GeneScoreMatrix",
                                 matrixName = "GeneIntegrationMatrix",
                                 reducedDims = "IterativeLSI",
                                 seRNA = rna_so,
                                 groupList = groupList,
                                 groupRNA = "celltype",
                                 nameCell = "co_pred_type",
                                 nameGroup = "co_pred_group",
                                 nameScore = "co_pred_score",
                                 force = TRUE
)
```

#### Constrained confusion matrix
```{r}
co_conf_mat <- as.matrix(confusionMatrix(proj$Clusters, proj$co_pred_group))
heatmap(co_conf_mat)
```

#### Map labels
```{r}
old_labels <- rownames(co_conf_mat)
new_labels <- colnames(co_conf_mat)[apply(co_conf_mat, 1, which.max)]

label_map <- mapLabels(old_labels,
                       oldLabels = old_labels,
                       newLabels = new_labels)
label_map
```

#### Transfer labels to ATAC
```{r}
proj$seurat_pbmc_type <- mapLabels(proj$Clusters, 
                                   newLabels = label_map, 
                                   oldLabels = old_labels)
```

[Return to Contents](#contents)

<a id="archr_output"></a>

#### Write results as a csv
```{r}
results <- data.frame(barcodes = sub(".+#","",rownames(proj@cellColData)),
                      doublet_score = proj@cellColData$DoubletScore,
                      doublet_enrichment = proj@cellColData$DoubletEnrichment,
                      seurat_pbmc_type = proj$seurat_pbmc_type,
                      seurat_pbmc_type_score = proj$co_pred_score)

out_table_path <- file.path(out_dir,
                            paste0(in_sample, "_archr_results.csv.gz"))
fwrite(results,
       out_table_path)
```

```{r}
file.remove(temp_dir,
            recursive = TRUE)
temp_arrows <- list.files("./", pattern = ".arrow")
file.remove(temp_arrows)
```

[Return to Contents](#contents)

<a id="json_out"></a>

### Write QC JSON

```{r Save QC JSON}
stm(paste0("No JSON output yet"))
# 
# qc_list_json <- jsonlite::toJSON(qc_list,
#                                  auto_unbox = TRUE,
#                                  pretty = TRUE)
# 
# writeLines(qc_list_json,
#            out_json)
```

[Return to Contents](#contents)

<a id="session_info"></a>

## Session Information

```{r Session Info}
sessionInfo()
```

Total time elapsed
```{r Show Time}
end_time <- Sys.time()
diff_time <- end_time - start_time
time_message <- paste0("Elapsed Time: ", 
                       round(diff_time, 3),
                       " ", units(diff_time))
print(time_message)
stm(time_message)
stm("ArchR analysis process complete.")
```

[Return to Contents](#contents)
