---
title: "Figure 3 ICICLE"
author: "Lucas Graybuck"
date: "5/29/2020"
output: html_document
---

## Setup

Load Packages
```{r}
quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}

quiet_library(H5weaver)
quiet_library(ATAComb)
quiet_library(uwot)
quiet_library(readr)
quiet_library(dplyr)
quiet_library(purrr)
quiet_library(ggplot2)
quiet_library(ggrastr)
quiet_library(cowplot)
quiet_library(scrattch.vis)
quiet_library(dendextend)
options(stringsAsFactors = FALSE)
set.seed(42)
```

Source directories
```{r}
filt_meta_dir <- "../../Data/X037_icicle/filtered_metadata/"
archr_res_dir <- "../../Data/X037_icicle/archr_results/"
peak_mat_dir <- "../../Data/X037_icicle/peaks_h5/"
adt_count_dir <- "../../Data/X037_icicle/adt_counts/"
```

Selected wells:  
X037-AP0C1W1, X037_icicle-1, Well 1 of ICICLE-seq Expt. 7  
X037-AP0C1W2, X037_icicle-2, Well 2 of ICICLE-seq Expt. 7  
```{r}
wells_to_process <- list(icicle_1 = "X037_icicle-1",
                         icicle_1 = "X037_icicle-2")
well_labels <- c("ICICLE: Well 1",
                 "ICICLE: Well 2")
```

Locate files
```{r}
filt_meta_files <- map2(filt_meta_dir, wells_to_process, list.files, full.names = TRUE)
archr_res_files <- map2(archr_res_dir, wells_to_process, list.files, full.names = TRUE)
peak_mat_files <- map2(peak_mat_dir, wells_to_process, list.files, full.names = TRUE)
adt_count_files <- map2(adt_count_dir, wells_to_process, list.files, full.names = TRUE)
```

Display inputs
```{r}
print("Filtered Metadata Files:")
print(unlist(filt_meta_files))
```
```
print("ArchR Results Files:")
print(unlist(archr_res_files))
```

```{r}
print("Peak Matrix Files:")
print(unlist(peak_mat_files))
```

```{r}
print("ADT Count Files:")
print(unlist(adt_count_files))
```

Read metadata files
```{r message=FALSE}
filt_meta_list <- map(filt_meta_files, 
                      read_csv)
meta_list <- filt_meta_list
meta <- do.call(rbind, meta_list)
```

Read ArchR results files
```
archr_res_list <- map(archr_res_files, 
                      read_csv)
```

Read peak matrices
```{r}
peak_mats <- map(peak_mat_files,
                 read_h5_dgCMatrix)
```

Read ADT files
```{r message = FALSE}
adt_count_list <- map(adt_count_files, 
                      read_csv)
```

### Filtering to remove doublets
ArchR results are filtered to remove doublet cells.
```
meta_list <- map2(filt_meta_list, archr_res_list,
                  function(filt_meta, archr_res) {
                    left_join(archr_res, filt_meta)
                  })
celltypes <- unique(unlist(map(meta_list, function(meta) unique(meta$seurat_pbmc_type))))
celltype_colors <- data.frame(seurat_pbmc_type = celltypes,
                              seurat_pbmc_type_color = immutils::varibow(length(celltypes)))
#color bug
celltype_colors$seurat_pbmc_type_color[celltype_colors$seurat_pbmc_type == "Mono.CD16"] <- "#267DFF"
meta_list <- map(meta_list,
                  function(meta) {
                    left_join(meta, celltype_colors)
                  })
```

Filter ADT counts
```{r}
adt_count_list <- map2(adt_count_list, meta_list,
                       function(adt_count, meta) {
                         names(adt_count)[1] <- "original_barcodes"
                         names(adt_count) <- gsub("-",".",names(adt_count))
                         adt_count <- adt_count %>%
                           filter(original_barcodes %in% meta$original_barcodes)
                         adt_count$barcodes <- meta$barcodes[match(adt_count$original_barcodes, meta$original_barcodes)]
                         adt_count <- adt_count[match(meta$barcodes, adt_count$barcodes),]
                         adt_count
                       })
```


Filter peak data too
```{r}
peak_mats <- map2(peak_mats, meta_list,
                  function(peak_mat, meta) {
                    peak_mat[,meta$barcodes]
                  })

peak_mat <- do.call(cbind, peak_mats)
```


### Dimension reduction
```{r}
peak_lsi <- atac_lsi(peak_mat,
                     site_frequency_threshold = 0.001)
peak_pcs <- peak_lsi$pcs[,2:20]
```

```{r}
peak_umap <- umap(peak_pcs,
                   scale = FALSE,
                   min_dist = 0.2)
peak_umap <- as.data.frame(peak_umap)
names(peak_umap) <- c("umap_1","umap_2")

peak_umap_df <- cbind(meta, peak_umap)
```

```{r}
peak_plot_umap_log_unique <- ggplot(peak_umap_df) +
  geom_point_rast(aes(x = umap_1,
                      y = umap_2,
                      color = peaks_frac),
                  size = 0.2) +
  scale_color_viridis_c("FRIP",
                        limits = c(0,1)) +
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank()) +
  guides(color = guide_colourbar(title.position = "top",
                                 barheight = unit(0.1, "in")))

```

```{r fig.width = 2, fig.height = 2.25}
peak_plot_umap_log_unique
```


```
peak_plots_umap_celltype <- map2(peak_umap_dfs, well_labels,
                       function(peak_umap_df, well_label) {
                         ggplot(peak_umap_df) +
                           geom_point_rast(aes(x = umap_1,
                                               y = umap_2,
                                               color = seurat_pbmc_type_color)) +
                           theme_bw(base_size = 7) +
                           scale_color_identity() +
                           theme(legend.position = "bottom",
                                 axis.text = element_text(color = "black"),
                                 axis.ticks = element_blank(),
                                 panel.grid.minor = element_blank(),
                                 panel.border = element_blank()) +
                           ggtitle(well_label)
                       })
```

```{ fig.width = 7.5, fig.height = 2}
peak_plots_umap_celltype_grid
```

```{ fig.width = 1.5, fig.height = 2}
ggplot(celltype_colors) +
  geom_tile(aes(x = 1,
                y = nrow(celltype_colors):1,
                fill = seurat_pbmc_type_color)) +
  geom_text(aes(x = 1.6,
                y = nrow(celltype_colors):1,
                label = seurat_pbmc_type),
            hjust = 0,
            size = 3) +
  scale_fill_identity() +
  xlim(0,5)+
  theme_void()

```

```
ggsave("Figure_2A_umaps_celltype.pdf",
       peak_plots_umap_celltype_grid,
       width = 7.5, height = 2.25,
       useDingbats = FALSE)
```

### ADT dim reduction

```{r}
adt_features <- names(adt_count_list[[1]])[grepl("^CD|^HLA", names(adt_count_list[[1]]))]

adt_mat_list <- map(adt_count_list,
                    function(adt_count) {
                      adt_mat <- as.matrix(adt_count[, adt_features])
                      rownames(adt_mat) <- adt_count$barcodes
                      t(adt_mat)
                    })

adt_mat <- do.call(cbind, adt_mat_list)
```

```{r}
norm_adt_mat <- adt_mat / colSums(adt_mat)[col(adt_mat)] * 1e3
norm_adt_mat <- log10(norm_adt_mat + 1)
norm_adt_mat[is.na(norm_adt_mat)] <- 0
```

```{r}
adt_umap <- umap(t(norm_adt_mat),
                 min_dist = 0.2)
adt_umap <- as.data.frame(adt_umap)
names(adt_umap) <- c("adt_umap_1","adt_umap_2")

adt_umap_df <- cbind(peak_umap_df, adt_umap)
adt_umap_df <- left_join(adt_umap_df, 
                         do.call(rbind,adt_count_list))
```

```{r}
adt_plot_umap_frip <- ggplot(adt_umap_df) +
  geom_point_rast(aes(x = adt_umap_1,
                      y = adt_umap_2,
                      color = peaks_frac),
                  size = 0.2) +
  scale_color_viridis_c("FRIP",
                        limits = c(0,1)) +
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank()) +
  guides(color = guide_colourbar(title.position = "top",
                                 barheight = unit(0.1, "in")))

```

```{r fig.width = 2, fig.height = 2.25}
adt_plot_umap_frip
```

```{r}
adt_plot_umap_logtotal <- ggplot(adt_umap_df) +
  geom_point_rast(aes(x = adt_umap_1,
                      y = adt_umap_2,
                      color = log10(total + 1)),
                  size = 0.2) +
  scale_color_viridis_c(bquote(log[10]("Unique ADTs + 1")),
                        limits = c(2,4.5)) +
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank()) +
  guides(color = guide_colourbar(title.position = "top",
                                 barheight = unit(0.1, "in")))

```

```{r fig.width = 2, fig.height = 2.25}
adt_plot_umap_logtotal
```

```{r}
adt_plot_umap_log10_features <- map(adt_features,
                                    function(feat) {
                                      feat <- rlang::parse_expr(feat)
                                      feat_range <- range(adt_umap_df[[feat]])
                                      log_feat_range <- log10(feat_range + 1)
                                      
                                      ggplot(adt_umap_df) +
                                        geom_point_rast(aes(x = adt_umap_1,
                                                            y = adt_umap_2,
                                                            color = log10(!!feat + 1)),
                                                        size = 0.5) +
                                        scale_color_viridis_c(feat,
                                                              limits = c(1, log_feat_range[2]),
                                                              na.value = "#000000") +
                                        theme_bw(base_size = 7) +
                                        theme(legend.position = "bottom",
                                              axis.text = element_text(color = "black"),
                                              axis.ticks = element_blank(),
                                              panel.grid.minor = element_blank(),
                                              panel.border = element_blank()) +
                                        guides(color = guide_colourbar(title.position = "top",
                                                                       barheight = unit(0.1, "in")))
                                    })
```

```{r}
adt_plot_umap_log10_features_grid <- plot_grid(
  plotlist = adt_plot_umap_log10_features,
  ncol = 4
)
```

```{r fig.width = 6, fig.height = 6}
adt_plot_umap_log10_features_grid
```


```{r}
peak_plot_umap_log10_features <- map(adt_features,
                                    function(feat) {
                                      feat <- rlang::parse_expr(feat)
                                      feat_range <- range(adt_umap_df[[feat]])
                                      log_feat_range <- log10(feat_range + 1)
                                      
                                      ggplot(adt_umap_df) +
                                        geom_point_rast(aes(x = umap_1,
                                                            y = umap_2,
                                                            color = log10(!!feat + 1)),
                                                        size = 0.5) +
                                        scale_color_viridis_c(feat,
                                                              limits = c(1, log_feat_range[2]),
                                                              na.value = "#000000") +
                                        theme_bw(base_size = 7) +
                                        theme(legend.position = "bottom",
                                              axis.text = element_text(color = "black"),
                                              axis.ticks = element_blank(),
                                              panel.grid.minor = element_blank(),
                                              panel.border = element_blank()) +
                                        guides(color = guide_colourbar(title.position = "top",
                                                                       barheight = unit(0.1, "in")))
                                    })
```

```{r}
peak_plot_umap_log10_features_grid <- plot_grid(
  plotlist = peak_plot_umap_log10_features,
  ncol = 4
)
```

```{r fig.width = 6, fig.height = 6}
peak_plot_umap_log10_features_grid
```


### UMAP Coordinate color cross-mapping

```{r}
xy2hex <- function(x, y) 
{
  x <- (x - min(x)) / max(x - min(x)) - 0.5
  y <- (y - min(y)) / max(y - min(y)) - 0.5
  h <- atan(y / x) * 2 / pi
  h <- (h - min(h)) / max(h - min(h))
  v <- sqrt(x^2 + y^2)
  v <- (v - min(v)) / max(v - min(v))
  v <- v / 2 + 0.5
  
  hsv(h = h, v = v, s = 0.8)
}

peak_umap_hex <- xy2hex(adt_umap_df$umap_1,
                        adt_umap_df$umap_2)

adt_umap_df$peak_umap_hex <- peak_umap_hex
```

```{r fig.width = 4, fig.height = 2}
p1 <- ggplot(adt_umap_df) +
  geom_point_rast(aes(x = umap_1,
                      y = umap_2,
                      color = peak_umap_hex),
                  size = 0.5) +
  scale_color_identity() + 
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())

p2 <- ggplot(adt_umap_df) +
  geom_point_rast(aes(x = adt_umap_1,
                      y = adt_umap_2,
                      color = peak_umap_hex),
                  size = 0.5) +
  scale_color_identity() + 
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())
plot_grid(p1, p2, ncol = 2)
```
### Cluster in ATAC space

Perform Jaccard-Louvain clustering
```{r}
peak_clusters <- atac_clustering(peak_pcs,
                                 k = 9)
adt_umap_df$peak_cluster <- peak_clusters$cl
```

Separate data from annotations for use with scrattch.vis
```{r}
data_df <- adt_umap_df
data_df$sample_name <- data_df$barcodes
data_df <- data_df %>% 
  select(sample_name, all_of(adt_features))
data_df[is.na(data_df)] <- 0

anno_df <- adt_umap_df %>%
  mutate(sample_name = barcodes) %>%
  scrattch.io::annotate_cat(peak_cluster) %>%
  mutate(peak_cluster_color = ifelse(nchar(peak_cluster_color) == 5,
                                     paste0(peak_cluster_color, "FF"),
                                     peak_cluster_color)) %>%
  select(sample_name, everything()) %>%
  select(-all_of(adt_features)) %>%
  mutate(peak_cluster_id = as.numeric(peak_cluster_label))
```

Compute medians of data columns for use in cluster hierarchy and ordering
```{r fig.width = 4, fig.height = 3}
adt_scaled <- scale_gene_data(data_df,
                              genes = adt_features,
                              scale_type = "log10")
adt_medians <- group_stats(adt_scaled,
                           value_cols = adt_features,
                           anno_df,
                           grouping = "peak_cluster_id",
                           stat = "mean")
adt_medians_mat <- as.matrix(adt_medians[,-1])
rownames(adt_medians_mat) <- adt_medians$peak_cluster_id

adt_medians_dist <- dist(adt_medians_mat)
adt_medians_hc <- hclust(adt_medians_dist,
                         method = "ward.D2")
cluster_order <- rownames(adt_medians_mat)[adt_medians_hc$order]
plot(adt_medians_hc)
```

Rotate for plots
```{r fig.width = 4, fig.height = 3}
adt_medians_dend <- as.dendrogram(adt_medians_hc)
adt_medians_dend <- adt_medians_dend %>%
  rotate(as.character(c(8,4,7,16,3,13,2,6,9,15,11,12,5,14,10,1)))
plot(adt_medians_dend)
```

Update cluster IDs to set order
```{r}
anno_df <- anno_df %>%
  mutate(peak_cluster_id = match(peak_cluster_label, labels(adt_medians_dend)))
```

```{r fig.width = 2, fig.height = 4}
p1 <- ggplot() +
  geom_point_rast(data = anno_df,
                  aes(x = umap_1,
                      y = umap_2,
                      color = peak_cluster_color),
                  size = 0.5) +
  geom_text(data = anno_df %>%
              group_by(peak_cluster_label) %>%
              summarise(umap_1 = mean(umap_1),
                        umap_2 = mean(umap_2)),
            aes(x = umap_1,
                y = umap_2,
                label = peak_cluster_label)) +
  scale_color_identity() +
  theme_bw(base_size = 7) +
  theme(legend.position = "none",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())

p2 <- ggplot() +
  geom_point_rast(data = anno_df,
                  aes(x = adt_umap_1,
                      y = adt_umap_2,
                      color = peak_cluster_color),
                  size = 0.5) +
  # geom_text(data = anno_df %>%
  #             group_by(peak_cluster_label) %>%
  #             summarise(adt_umap_1 = mean(adt_umap_1),
  #                       adt_umap_2 = mean(adt_umap_2)),
  #           aes(x = adt_umap_1,
  #               y = adt_umap_2,
  #               label = peak_cluster_label)) +
  scale_color_identity() +
  theme_bw(base_size = 7) +
  theme(legend.position = "none",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())

plot_grid(p1, p2, ncol = 1)
```

```{r fig.width = 6, fig.height = 6}
library(scrattch.vis)
group_violin_plot(data = data_df,
                  anno = anno_df,
                  genes = adt_features,
                  grouping = "peak_cluster",
                  log_scale = TRUE)
```

```{r fig.width = 4, fig.height = 4}
library(scrattch.vis)
sample_fire_plot(data = data_df,
                 anno = anno_df,
                 genes = adt_features,
                 grouping = "peak_cluster",
                 log_scale = TRUE,
                 label_height = 10,
                 max_width = 20)

```
### Joint reduction and clustering
```{r}
peak_pcs_with_adts <- cbind(peak_pcs[,1:11], t(norm_adt_mat))

joint_umap <- umap(peak_pcs_with_adts,
                   min_dist = 0.2,
                   scale = "scale")

anno_df$joint_umap_1 <- joint_umap[,1]
anno_df$joint_umap_2 <- joint_umap[,2]
```

```{r fig.width = 2, fig.height = 4}
p1 <- ggplot() +
  geom_point_rast(data = anno_df,
                  aes(x = umap_1,
                      y = umap_2,
                      color = peak_cluster_color),
                  size = 0.5) +
  geom_text(data = anno_df %>%
              group_by(peak_cluster_label) %>%
              summarise(umap_1 = mean(umap_1),
                        umap_2 = mean(umap_2)),
            aes(x = umap_1,
                y = umap_2,
                label = peak_cluster_label)) +
  scale_color_identity() +
  theme_bw(base_size = 7) +
  theme(legend.position = "none",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())

p2 <- ggplot() +
  geom_point_rast(data = anno_df,
                  aes(x = joint_umap_1,
                      y = joint_umap_2,
                      color = peak_cluster_color),
                  size = 0.5) +
  # geom_text(data = anno_df %>%
  #             group_by(peak_cluster_label) %>%
  #             summarise(adt_umap_1 = mean(adt_umap_1),
  #                       adt_umap_2 = mean(adt_umap_2)),
  #           aes(x = adt_umap_1,
  #               y = adt_umap_2,
  #               label = peak_cluster_label)) +
  scale_color_identity() +
  theme_bw(base_size = 7) +
  theme(legend.position = "none",
        axis.text = element_text(color = "black"),
        axis.ticks = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())

plot_grid(p1, p2, ncol = 1)
```


```{r}
sessionInfo()
```