---
title: "Data retrieval from GEO"
output: html_notebook
---

In this RMarkdown notebook, we'll demonstrate how to download the supplementary files from our GEO entries, which can be used for downstream analysis of our data.


We'll use the `GEOquery` BioConductor package, as well as the `purrr` package for help with handling multiple samples, and `gt` to make some nicely formatted tables.

```{r}
library(GEOquery)
library(purrr)
library(gt)
```


First, let's find our Series accession in GEO. The series for Swanson, et al. 2021 is "GSE158013":

```{r}
series_info <- getGEO("GSE158013",
                      GSEMatrix = FALSE)
```

Next, we'll extract the sample metadata:

```{r}
sample_info <- GSMList(series_info)
```

A lot of information about these samples is stored in the "characteristics" fields in the sample GEO metadata. Let's use purrr's `map_dfr()` to parse this data and return a nice data.frame to work with:

```{r}
sample_characteristics <- map_dfr(
  sample_info,
  function(si) {
    raw_chars <- Meta(si)$characteristics_ch1
    char_names <- c("geo_accession",sub(":.+","", raw_chars))
    char_vals <- c(Meta(si)$geo_accession,sub(".+: ","", raw_chars))
    names(char_vals) <- char_names
    as.data.frame(as.list(char_vals))
  }
)

sample_characteristics %>%
  gt(rowname_col = "geo_accession",
     groupname_col = "library_type") %>%
  tab_options(
    table.font.size = 10,
    column_labels.font.weight = "bold",
    row_group.font.size = 12,
    row_group.font.weight = "bold"
  )
```

Let's save a copy to .csv for later use:
```{r}
write.csv(sample_characteristics,
          "GEO_sample_characteristics.csv",
          row.names = FALSE)
```


Now, we can use the accession ids and metadata to download relevant files.

First, let's make a directory for each library type:

```{r}
library_types <- unique(sample_characteristics$library_type)

walk(library_types, dir.create)
```

Let's make a nice table of supplementary files to retrieve (should be useful for later analysis):

```{r}
supplementary_files <- map_dfr(
  sample_info,
  function(si) {
    si_meta <- Meta(si)
    sf <- si_meta[grepl("supplementary_file", names(si_meta))]
    sf <- unlist(sf)
    names(sf) <- NULL
    data.frame(geo_accession = si_meta$geo_accession,
               supplementary_file = sf)
  }
)

head(supplementary_files) %>%
  gt() %>%
  tab_options(
    table.font.size = 10,
    column_labels.font.weight = "bold"
  )
```

```{r}
write.csv(supplementary_files,
          "GEO_supplementary_files.csv",
          row.names = FALSE)
```

Then, we can download all of the GEO supplementary files for each sample:
```{r}
walk(
  library_types,
  function(library_type) {
    type_samples <- sample_characteristics[sample_characteristics$library_type == library_type,]
    sample_files <- supplementary_files[supplementary_files$geo_accession %in% type_samples$geo_accession,]
    
    walk(sample_files$supplementary_file,
         function(sf) {
           target <- file.path(library_type,
                               sub(".+/","",sf))
           download.file(sf, target)
         })
  }
  )
```

```{r}
sessionInfo()
```

